New method to describe the zonal symmetries and asymmetries of the Southern Annular Mode

### 2020-09-23 analysis/paper/paper.Rmd
casi casi

---
twocol: TRUE  #  Use FALSE for submission. 
title: Title here
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro Díaz
affiliation: "CIMA UBA blablabla"
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

@@

yada yada SAM yada yada circulation.. yada yada so important. yada yada many impacts.

@fogt2012 studied the characteristics of the asymmetric structure of the SAM. It computed the zonally anomalous component of mean sea level pressure (MSLP) composites for positive and negative SAM events and created two indices by projecting MSLP fields ontno them. However, the use of composites leads to some issues that affects the interpretability of the results. First, they can be dependent on the choice of threshold used to define positive and negative events. Secondly, by discarting data that don't meet the threshold, they don't use all the information avaiable. Due to the realtive short timeframe used, this leads to some composites being composed of as little as 4 years. Third, the resulting composites corresponding to each polarity and season are derived from the average of different amount of fields and from different years. This last issue is particularly important in light of the changing structure of the SAM before and after 1980 [@silvestri2009]. In @fogt2012, the DJF SAM+ composite uses only 7 years, 5 of which are later than 1988, whereas all of the 8 years used for their DJF SAM- composite are from earlier than 1988.

@@

To separate between the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa. The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the "Full", "Asymmetric" and "Symmetric" indices, by regressing each geopotential field on the respective patterns (weighting by the cosine of latitude). The three indices are normalised by dividing them by the standard deviation of the "Full" index at each level. As a result, the magnitude between indices is comparable. However, only "Full" index will have unit standard deviation per definition.

```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height. Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right). Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
  setnames("hgt", "full") %>% 
  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]

@@

```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM and density estimates.", fig.width=6, fig.height=3, fig.env="figure*"}

indices %>% 
  .[lev %in% c(50, 700) & term != "full"] %>% 
  ggplot(aes(estimate_norm)) +
  geom_density(aes(color = term)) +
  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam, guide = "none") +
  scale_y_continuous(NULL, trans = scales::reverse_trans(), labels = NULL, breaks = NULL) +
  scale_x_continuous(NULL, labels = NULL, breaks = NULL) +
  coord_flip()  +
  theme(strip.text = element_blank()) +

  indices %>% 
  .[lev %in% c(50, 700) & term != "full"] %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
  ggplot(aes(time, estimate_norm)) +
  geom_line(aes(color = term)) +
  # ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
  #                           aes(color = term, width = ..density.., x = min(time - 1e7))) +
  # geom_smooth(aes(color = term), method = "loess", n = 481, span = 10*12/481, method.args = list(degree = 1)) +
  # geom_point(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
  #                         time == as.Date("1987-08-01") & term == "asym"]) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
  scale_y_continuous(NULL) +
  # scale_x_datetime(NULL, expand = c(0, 0)) +
  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
  tag_facets() +
  plot_layout(widths = c(1, 4))
```

@@

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987). We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. As a consequence, statospheric Asymmetric index tend to precede corresponding Symmetric index.

```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level for lag zero and lag -1 (Asymmetric leads Symmetric).", fig.width=3, fig.height=3}
cors %>% 
  .[lag %in% c(0, -1)] %>% 
  ggplot(aes(lev, acf)) +
  geom_line(aes(color = factor(lag, levels = c(0, -1), ordered = TRUE),
                linetype = factor(lag, levels = c(0, -1), ordered = TRUE))) +
  scale_x_level() +
  scale_y_continuous("Correlation", limits = c(0, 1)) + 
  scale_linetype_manual(NULL, values = c(1, 5), labels = c("0" = "Zero-lag", 
                                                           "-1" = "Asymmetric leads\nSymmetric (1 month)")) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c("0" = "Zero-lag", 
                                                         "-1" = "Asymmetric leads\nSymmetric (1 month)")) +
  coord_flip()
```

```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM.", fig.width = 6, fig.height=3, fig.env="figure*"}
tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
  as.data.table() %>% 
  setnames(., colnames(.), c("item1", "item2"))

@@

```{r trends, fig.cap = "Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
annual_trend <- indices %>%
  copy() %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  .[, season := "Annual"]

indices %>% 
  copy() %>% 
  # .[time > as.Date("1990-01-01")] %>%
  .[, season := season(time)] %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  rbind(annual_trend) %>% 
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  setnames("type", "term") %>% 
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
              fill = "#95a3ab", color = "black", 
              alpha = .6, size = 0.1) +
  geom_line() +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  geom_hline(yintercept = 0) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("cr") 
```

@@

```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends for explained variance of each index at each level for the period 1979-2018. Shading indicates the 95\\% confidence interval."}
indices %>% 
  .[term != "full"] %>% 
  .[, FitLm(r.squared, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>%
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  ggplot(aes(lev, estimate)) +
  geom_hline(yintercept = 0) +
  geom_ribbon(aes(fill = type, color = type, 
                  ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error), 
              size = 0.2, alpha = 0.2) +
  geom_line(aes(color = type)) +
  scale_x_level() +
  scale_y_continuous("Trend in R^2") +
  scale_fill_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill"), labels = lab_sam) +
  coord_flip() 
```

@@

```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef. Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>%
  .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, exclude =  0),
                      size = 0.3,
                      global.breaks = FALSE) +
  # geom_contour2(aes(z = env)) +
  geom_qmap(~.x[lat <= -20]) +
  geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
                               lat = -49, term = "asym", lev = 700), 
             shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-45, 45),
                       oob = scales::squish,
                       breaks = AnchorBreaks(0, 10, 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr") +
  axis_labs_smol  +
  no_grid
```

@@

```{r wave-amplitude, fig.cap = "Planteray wave amplitude for the regression patterns at 50 and 700 hPa. Note the varying x axis.", fig.width=3, fig.height=3}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  scale_color_brewer("Wave-number", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free") +
  tag_facets("cr", position = "tr")  +
  no_grid +
  theme(strip.text = element_text(size = 7))
```

@@

vertical_regr <- ReadNetCDF(ERA5(), 
                            subset = list(time = c("1979-01-01", "2018-12-31"),
                                          latitude = c(-65, -40)),
                            vars = c(hgt = "z")) %>% 
  normalise_coords() %>% 
  .[, .(hgt = weighted.mean(hgt, w = cos(lat*pi/180))), by = .(lon, lev, time)] %>% 
  .[, hgt := hgt/9.8] %>% 
  .[, hgt_a := hgt - mean(hgt), by = .(lon, lev, month(time))] %>% 
  indices_for_regr[., on = .NATURAL, allow.cartesian = TRUE] %>% 
  .[, combine_lists(FitLm(hgt_a, full, se = TRUE),
                    FitLm(hgt_a, sym, asym, se = TRUE)),
    by = .(lon, lev, index_level)] %>% 
  rm_intercept() %>% 
  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, index_level, lev)] %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
```

```{r vertical-regression, fig.cap = "Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South. (\\#FIXME this caption needs some love)", fig.width = 3, fig.height=3}
vertical_regr %>% 
  .[index_level %in% c(50, 700)] %>% 
  .[term == "asym"] %>%
  ggplot(aes(lon, lev)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, 10, 0)) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, 10, 0),
                      size = 0.3,
                      global.breaks = FALSE) +
  # geom_contour2(aes(z = p.val), breaks = 0.01) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(13), 
                       oob = scales::squish,
                       breaks = MakeBreaks(binwidth = 10, exclude = 0)) +
  scale_y_level() +
  scale_x_longitude(ticks = 60)  +
  facet_grid(index_level~., labeller = labeller(index_level = lev.lab)) +
  tag_facets(position = "bl") 
```

@@

```{r}
#  enso %>% 
#   .[indices, on = "time", allow.cartesian = TRUE] %>% 
#   .[, with(cor.test(estimate, value), 
#            list(cor = estimate, 
#                 low = conf.int[1], 
#                 upp = conf.int[2])), by = .(lev, term, index, season(time))] %>% 
#   .[index == "oni"] %>% 
# ggplot(aes(lev, cor)) +
#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
#   geom_line(aes(color = term)) +
#   scale_x_level() +
#   coord_flip() +
#   facet_wrap(~season)
# 
# enso %>% 
#   .[indices_wide, on = "time", allow.cartesian = TRUE] %>% 
#   .[, partial_cor(value, asym, sym), by = .(index, lev, season(time))] %>% 
#   .[index == "oni"] %>%
# ggplot(aes(lev, partial_correlation)) +
#   # geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
#   geom_line(aes(color = term, group = interaction(term, index))) +
#   scale_x_level() +
#   coord_flip() +
#   facet_wrap(~season)
```

>>>>

---
twocol: FALSE  #  Use FALSE for submission. 
title: New method to describe the zonal symmetries and asymmetries of the Southern Annular Mode
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro Díaz
affiliation: |
  Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
  CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
  CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

@@

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a; fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020]. Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.

However, computed as the leading Empirical Orthogonal Function (EOF) of Sea Level Pressure or low-level geopotential height, the SAM spatial structure contains noticeable deviations from this zonally symmetric description, particularly in the Pacific Ocean region. These zonal asymmetries are not widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009]. At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availabilty of dense observations in the Southern Hemisphere [@jones2009].

We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012. However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988). This is particularly important due to the inhomogeneities in renalysis products prior to the sattelite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].

Our objective is, then, to systematically characterise the zonally asymmetric component of the SAM variability by constructing two indices which aim to capture exclusively the variability of the symmetric and asymmetric component each. We then analyse their temporal variability, trends and vertical coherence. We study the spatial patterns described by the variability exclusive to each index. Finally, we investigate theri relationship with temperature and recipitation anomalies.

@@

To separate between the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa. The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression ofeach monthly geopotential field on the respective patterns (weighting by the cosine of latitude). The three indices are normalised by dividing them by the standard deviation of the "Full" index at each level. As a result, the magnitude between indices is comparable. However, only "Full" index will have unit standard deviation per definition. From the regression, we also use the explained variance of each pattern as a indication of the degere of symmetry or asymmetry of each monthly field.

```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
  setnames("hgt", "full") %>% 
  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]

@@

```{r asymsam-timeseries, fig.cap = "Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa. To the left, probability density estimate of each index. Series are standarised by the standard deviation of the Full SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
indices %>% 
  .[lev %in% c(50, 700) & term != "full"] %>% 
  ggplot(aes(estimate_norm)) +
  geom_density(aes(color = term)) +
  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam, guide = "none") +
  scale_y_continuous(NULL, trans = scales::reverse_trans(), labels = NULL, breaks = NULL) +
  scale_x_continuous(NULL, labels = NULL, breaks = NULL) +
  coord_flip()  +
  theme(strip.text = element_blank()) +

  indices %>% 
  .[lev %in% c(50, 700) & term != "full"] %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
  ggplot(aes(time, estimate_norm)) +
  geom_line(aes(color = term)) +
  # ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
  #                           aes(color = term, width = ..density.., x = min(time - 1e7))) +
  # geom_smooth(aes(color = term), method = "loess", n = 481, span = 10*12/481, method.args = list(degree = 1)) +
  # geom_point(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
  #                         time == as.Date("1987-08-01") & term == "asym"]) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
  scale_y_continuous(NULL) +
  scale_x_datetime(NULL, expand = c(0, 0)) +
  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
  tag_facets() +
  plot_layout(widths = c(1, 4))
```

@@

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987). We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. As a consequence, statospheric Symmetric index tend to precede corresponding Asymmetric index.

```{r cor-lev, fig.cap = "Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric).", fig.width=3, fig.height=3}
cors %>% 
  .[lag %in% c(0, -1)] %>% 
  ggplot(aes(lev, acf)) +
  geom_line(aes(color = factor(lag, levels = c(0, -1), ordered = TRUE),
                linetype = factor(lag, levels = c(0, -1), ordered = TRUE))) +
  scale_x_level() +
  scale_y_continuous("Correlation", limits = c(0, 1)) + 
  scale_linetype_manual(NULL, values = c(1, 5), labels = c("0" = "Zero-lag", 
                                                           "-1" = "Symmetric leads\nAsymmetric by 1 month")) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c("0" = "Zero-lag", 
                                                         "-1" = "Symmetric leads\nAsymmetric by 1 month")) +
  coord_flip()
```

```{r cross-correlation, fig.cap = "Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM.", fig.width = 6, fig.height=3, fig.env="figure*"}
tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
  as.data.table() %>% 
  setnames(., colnames(.), c("item1", "item2"))

@@

```{r trends, fig.cap = "Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
annual_trend <- indices %>%
  copy() %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  .[, season := "Annual"]

indices %>% 
  copy() %>% 
  # .[time > as.Date("1990-01-01")] %>%
  .[, season := season(time)] %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  rbind(annual_trend) %>% 
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
  setnames("type", "term") %>% 
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_hline(yintercept = 0) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
              fill = "#95a3ab", color = "black", 
              alpha = .6, size = 0.1) +
  geom_line() +
  # stat_subset(aes(subset = pval <= 0.05), shape = 4, size = 2) +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("cr") 
```

@@

```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018. Shading indicates the 95\\% confidence interval."}
indices %>% 
  .[term != "full"] %>% 
  .[, FitLm(r.squared, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>%
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  ggplot(aes(lev, estimate)) +
  geom_hline(yintercept = 0) +
  geom_ribbon(aes(fill = type, color = type, 
                  ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error), 
              size = 0.2, alpha = 0.2) +
  geom_line(aes(color = type)) +
  scale_x_level() +
  scale_y_continuous("Change in explained variance per decade", label = scales::percent_format()) +
  scale_fill_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill"), labels = lab_sam) +
  coord_flip() 
```

@@

```{r 2d-regr, fig.cap = "Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef. Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index. ", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>%
  .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, exclude =  0),
                      size = 0.3,
                      global.breaks = FALSE) +
  # geom_contour2(aes(z = env)) +
  geom_qmap(~.x[lat <= -20]) +
  geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
                               lat = -49, term = "asym", lev = 700), 
             shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL,
                   minor_breaks = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-45, 45),
                       oob = scales::squish,
                       breaks = AnchorBreaks(0, 10, 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr") +
  axis_labs_smol  +
  no_grid +
  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
```

@@

```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean. Note the differnet x axis.", fig.width=3, fig.height=3}
indices_2d_yearly %>% 
  .[lat < -20] %>%
  .[lev %in% c(50, 700)] %>% 
  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  scale_color_brewer("Wave-number", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude(ticks = 20) +
  coord_flip() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free") +
  tag_facets("cr", position = "tr")  +
  no_grid +
  theme(strip.text = element_text(size = 7))
```

@@

vertical_regr <- ReadNetCDF(ERA5(), 
                            subset = list(time = c("1979-01-01", "2018-12-31"),
                                          latitude = c(-65, -40)),
                            vars = c(hgt = "z")) %>% 
  normalise_coords() %>% 
  .[, .(hgt = weighted.mean(hgt, w = cos(lat*pi/180))), by = .(lon, lev, time)] %>% 
  .[, hgt := hgt/9.8] %>% 
  .[, hgt_a := hgt - mean(hgt), by = .(lon, lev, month(time))] %>% 
  indices_for_regr[., on = .NATURAL, allow.cartesian = TRUE] %>% 
  .[, FitLm(hgt_a, asym, sym, se = TRUE),
    by = .(lon, lev, index_level)] %>% 
  rm_intercept() %>% 
  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, index_level, lev)] %>% 
  periodic(lon = c(0, 360)) %>% 
  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
```

```{r vertical-regression, fig.cap = "Regression between monthly geopotential anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM). (a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa.", fig.width = 3, fig.height=3}
vertical_regr %>% 
  .[index_level %in% c(50, 700)] %>% 
  .[term == "asym"] %>%
  ggplot(aes(lon, lev)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, 10, 0)) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, 10, 0),
                      size = 0.3,
                      global.breaks = FALSE) +
  # geom_contour2(aes(z = p.val), breaks = 0.01) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(13), 
                       oob = scales::squish,
                       breaks = MakeBreaks(binwidth = 10, exclude = 0)) +
  scale_y_level() +
  scale_x_longitude(ticks = 60)  +
  facet_grid(index_level~., labeller = labeller(index_level = lev.lab)) +
  tag_facets(position = "bl") 
```

@@

The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989]. Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO. The correlation between the full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[index == "oni"]$full`. Consistent with @fan2007, we show that this relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni"]$asym` with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni"]$sym`). We performed the same analysis using the Multivariate Enso Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.
